# MIMOSA 2

## README

We run the code in the `README` at `RGLab/MIMOSA2` on GitHub`.

```{r }
library(MIMOSA2)
library(data.table)
library(ggplot2)
```

```{r , error = TRUE}
set.seed(100)
s <- simulate_MIMOSA2(effect = 2e-4, baseline_stim_effect = 5e-5, baseline_background = 2e-5, bg_effect = 2e-5, phi = c(50000, 50000, 50000, 50000), P = 100, rng = c(120000, 150000))
pos_mat <- matrix(c(s$ns0, s$ns1, s$nu0, s$nu1), ncol = 4, byrow = FALSE)
tot_mat <- s$Ntot
freq_tbl <- (pos_mat / tot_mat * 1e2) |>
  as.data.frame()
cat_tbl <- tibble::tibble(truth = s$truth)
cat_tbl <- cat_tbl |>
  dplyr::mutate(
    responder_type = gsub("\\d", "", truth),
    responder_level = gsub("[a-zA-Z]", "", truth)
  )
sim_tbl <- cat_tbl |>
  dplyr::mutate(pid = as.character(1:100)) |>
  dplyr::bind_cols(freq_tbl) |>
  tidyr::pivot_longer(
    nu1:ns0,
    names_to = "stim_timepoint",
    values_to = "freq"
  ) |>
  dplyr::mutate(
    stim = ifelse(grepl("s", stim_timepoint), "stim", "uns"),
    timepoint = ifelse(grepl("0$", stim_timepoint), "pre", "post")
  ) |>
  dplyr::select(-stim_timepoint) |>
  dplyr::select(-truth) |>
  dplyr::select(pid, responder_type, responder_level, timepoint, stim, freq) |>
  dplyr::mutate(
    responder_level = ifelse(
      responder_level == "", "1", responder_level
      )
  )
sim_tbl_bs <- sim_tbl |>
  dplyr::group_by(
    pid, responder_type, responder_level, timepoint
  ) |>
  UtilsCytoRSV::subtract_background(
    grp = c("pid", "responder_type", "responder_level", "timepoint"),
    stim = "stim",
    resp = "freq",
    uns = "uns"
  ) |>
  dplyr::select(-stim) 
ggplot(
  sim_tbl_bs,
  aes(
    x = timepoint,
    y = freq,
    col = responder_level
  )
) +
  geom_hline(yintercept = 0) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "y") +
  geom_boxplot(outlier.size = -1) +
  facet_wrap(~responder_type) +
  labs(x = "Timepoint", y = "Background-subtracted frequency")

ggplot(
  sim_tbl|>
    dplyr::mutate(
      timepoint_stim = paste0(timepoint, "_", stim)
    ),
  aes(
    x = timepoint_stim,
    y = freq,
    col = responder_level
  )
)  +
  geom_hline(yintercept = 0) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "y") +
  geom_boxplot(outlier.size = -1) +
  facet_wrap(~responder_type) +
  labs(x = "Timepoint", y = "Frequency")
```

- Meaning of output from `s`
  - `Ntot`
    - Total T cell counts for stim or uns (`s` or `u`) at the first or second timepoint (`0` or `1`).
  - `ns0`, `ns1`
    - Positive cell counts in stimulated samples at the first (`ns0`) and second (`ns1`) timepoints
  - `nu0`, `nu1`
    - Positive cell counts in unstimulated samples at the first (`nu0`) and second (`nu1`) timepoints
  - `truth`
    - character vector, with one value per participant.
    - Possible values: `"R1"  "R2"  "R3"  "R4"  "NR1" "NR2" "NR3" "NSR"`
      - `R[1-4]`
      - `NR[1-3]`
      - `NSR`
      - Not clear what `NSR` is.
    - These must classify participants, so whether the participant is a responsender or not.
    - In the `README`, we see that `R` is a response (stim > uns), `NR` is a non-responder (stim == uns) and NSR is a non-responder but has high background (non-specific response: stim == uns >> 0).
      - That's not quite true, if you look at the simulated data.
        - It seems that `R` is someone who responded to the vaccine, so there is a response and it is larger at time 2 than time 1 in the stim but not the unstim.
        - `NR` is someone who did not respond to the vaccine and does not respond in general to the antigen. 
        - `NSR` is someone who responded to the antigen but not because of the vaccine. 
    - Looking at the plot of background-subtracted frequencies:
      - `NR` all don't have higher stims post vacc than unstims post vacc
      - `R` all respond, in varying manners (e.g. R3 also increases unstim a lot post-vaccination)
    - Looking at the plot of frequncies:
     - `NSR` have a non-specific increase in the positive count between pre- and post

The code below breaks, however:
```{r , error = TRUE}
fit <- with(s, MIMOSA2(Ntot = Ntot, ns1 = ns1, nu1 = nu1, ns0 = ns0, nu0 = nu0, tol = 1e-5))
```

## MIMOSA2 example

```{r , error = TRUE}
s = simulate_MIMOSA2();
R = MIMOSA2(Ntot=s$Ntot, ns1 = s$ns1, nu1 = s$nu1, nu0 = s$nu0, ns0 = s$ns0,maxit=10)
```

## Alternate dataset

```{r , error = TRUE}
n_tot <- data.frame(
  nu1 = rep(1e2, 20),
  nu0 = rep(1e2, 20),
  ns1 = rep(1e2, 20),
  ns0 = rep(1e2, 20)
) |>
  as.matrix()
ns1 <- rnorm(20, 20)
ns0 <- rnorm(10, 20)
nu1 <- rnorm(5, 20)
nu0 <- rnorm(5, 20)
MIMOSA2::MIMOSA2(
  Ntot = n_tot,
  ns1 = ns1,
  ns0 = ns0,
  nu1 = nu1,
  nu0 = nu0,
  tol = 1e-1
)
```

## Debugging MIMOSA2


```{r , eval = FALSE}
debugonce(MIMOSA2:::initialize)
fit <- with(s, MIMOSA2(Ntot = Ntot, ns1 = ns1, nu1 = nu1, ns0 = ns0, nu0 = nu0, tol = 1e-5))
est <- optimx(
  par = thetahat, fn = sumcll, method = "bobyqa",
  pi_est = pi_est, z = inds, Ntot = Ntot, ns1 = ns1, nu1 = nu1,
  ns0 = ns0, nu0 = nu0
)
```
